---
title: "Topic Modeling for Text Data: R Code"
author: "Monnie McGee"
date: "June 9, 2026"
output: html_document
embed_resources: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE)
```

This R Markdown file contains the R code extracted from the Beamer lecture slides on Topic Modeling. Chunks are set to `eval = FALSE` by default so the document can knit even before the needed data objects are created. Change `eval = TRUE` for chunks you want to run.

## R Exercise: Tokenize Text

```{r 01-r-exercise-tokenize-text}
library(tidyverse)
library(tidytext)
library(reshape2)

documents <- read.csv("AbstractExample.csv")

# documents: data frame with columns id and text
tokens <- documents %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words, by = "word")

word_counts <- tokens %>%
  count(id, word, sort = TRUE)

dtm <- word_counts %>%
  cast_dtm(document = id,
           term = word,
           value = n)
```

## Fit LDA in R

```{r 02-fit-lda-in-r}
library(topicmodels)

# Fit a topic model with 5 topics
lda_fit <- LDA(dtm, k = 5, control = list(seed = 1234))

# Display the 10 most probable words for each topic
terms(lda_fit, 10)
```

## Visualize Top Words by Topic

```{r 03-visualize-top-words-by-topic}
beta <- tidy(lda_fit, matrix = "beta")

beta %>%
  group_by(topic) %>%
  slice_max(beta, n = 10) %>%
  ungroup() %>%
  ggplot(aes(beta, reorder_within(term, beta, topic))) +
  geom_col() +
  facet_wrap(~ topic, scales = "free") +
  scale_y_reordered()

top_terms <- beta  %>% 
  group_by(topic) %>%
  slice_max(beta, n = 10) %>%
  arrange(topic, desc(beta))

top_terms
```

What would you name the topics?

Most representative documents
```{r lda_rep}

# Document-topic probabilities
gamma <- tidy(lda_fit, matrix = "gamma")
gamma

# By each topic
gamma %>%
  group_by(topic) %>%
  slice_max(gamma, n = 1)
```


```{r Choosing K for LDA}
# Compare several values of K

lda3 <- LDA(dtm, k = 3, control = list(seed = 1234))
lda5 <- LDA(dtm, k = 5, control = list(seed = 1234))

terms(lda3, 10)
terms(lda5, 10)

## Or perplexity - lower is better
k_values <- 2:5

perplexities <- sapply(k_values, function(k) {
  fit <- LDA(
    dtm,
    k = k,
    control = list(seed = 1234)
  )
  perplexity(fit)
})

plot(
  k_values,
  perplexities,
  type = "b",
  xlab = "Number of Topics (K)",
  ylab = "Perplexity"
)
```

Perplexity measures how well the model predicts unseen words; lower values indicate better predictive performance, but not necessarily more interpretable topics. Perplexity plays a role similar to $R^2$ in regression. It measures how well the model explains or predicts the observed data. However, just as $R^2$ tends to increase when we add more predictors, perplexity tends to decrease when we add more topics.

## Fit an STM in R

LDA ignores meta data on the text (covariates). Structural Topic Modeling is an advancement on LDA that is meant to remedy that problem.

```{r 04-fit-an-stm-in-r}
library(readr)
library(stm)

abstracts <- read_csv("AbstractExample.csv")

# Add document metadata (I made this up)
abstracts$field <- c(
  "policy", "health",
  "environment", "health",
  "environment", "policy",
  "statistics",
  "environment"
)

# Add made up metadata to abstracts
processed <- textProcessor(
  documents = abstracts$text,
  metadata  = abstracts
)

# Create one object with everything in it. Work from this.
out <- prepDocuments(
  processed$documents,
  processed$vocab,
  processed$meta
)


stm_fit <- stm(
  documents = out$documents,
  vocab = out$vocab,
  K = 5,
  prevalence = ~ field,
  data = out$meta,
  seed = 1234
)
```

`searchK()` does more than fit models. It also tries to compute diagnostics such as held-out likelihood. With only a small number of short documents, the held-out step can fail. Here is the code, but we won't use it for our toy example.

```{r Choosing K for STM} 
#stm_k_results <- searchK(
#  documents = out$documents,
#  vocab = out$vocab,
#  K = c(3, 5),
#  prevalence = ~ field,
#  data = out$meta,
#  seed = 1234
#)

#plot(stm_k_results)
```

## Get topics
```{r stm_topics}
stm_topics <- labelTopics(stm_fit)
stm_topics

# Topic proportions
topicCorr(stm_fit)
colMeans(stm_fit$theta)

# Most representative documents
top_doc <- order(stm_fit$theta[,1],
                 decreasing = TRUE)[1:5]

top_doc
abstracts$text[top_doc]
```

What would you name the topics?

## Estimate Effects of Metadata

How do the topics change as a function of the model covariates? STM allows a regression-like model where the response variable is topic prevalence and the predictor is document metadata.

```{r 05-estimate-effects-of-metadata}

effects <- estimateEffect(
  1:5 ~ field,
  stm_fit,
  meta = out$meta
)

summary(effects)
plot(effects, covariate = "field", topics = c(1, 2, 3))
```
